Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-06-12 limits/ms.tex
theory section largely finished

The lens equation:
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor:
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4GD_L}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass \citep[e.g.][]{1992grle.book.....S}. For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

@@

The trouble with solving the lens equation for the mass distribution is that the solutions are non-unique and degenerate -- many different solutions can give a statistically good reproduction of the data. To see why this is, it is instructive to simplify equation \ref{eqn:lens_equation} for a circularly symmetric lens: 
%
\begin{equation}
\vec\beta = \vec\theta - \frac{\theta_E^2}{|\vec\theta|^2}\vec\theta
\label{eqn:lens_equation_circ}
\end{equation}
%
where $\theta_E^2 = \frac{D_{LS}}{D_S D_L} \frac{4 GM}{c^2}$ is the Einstein radius.

\begin{itemize}
  \item Steepness
  \item Shape
  \item $H_0$
  \item Rings
  \item Cosmology
\end{itemize}

\subsection{The utility of stellar mass estimates}

\subsection{Adding stellar kinematic constraints}

@@

\subsection{A new lens modeling framework: \Glass}

\Glass\ is an evolution of the concepts employed by the free form modeling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, but a user can write their own in the input file.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, it is planned
to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can be have a higher resolution to capture steep models. 
\end{enumerate}
%
By default, the mass distribution of the lens is described by uniform grid. However,
in the central region of a lensing galaxy where the mass profile may rise steeply,
a higher resolution can be specified that only applies to the inner pixels. These
pixels are subdivided into a smaller grid.
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to model region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additional allows for additional analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is group then a potential can be added for
each of the galaxies.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modeling solution space. This algorithm was
described and tested in \cite{}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

@@

In addition to the lens model aspects, \Glass\ can also run post processing
routines to analyze each model as it is produced. We have developed a number of
routines for fitting analytic models to the lensing mass profiles. One in
particular deprojects the 2d profile assuming a Hernquist profile for the
light. By fixing the velocity anisotropy $\beta$ we can determine the velocity
dispersion of the model. Of course, the true velocity dispersion is known for
the mock galaxies and we can compare our method to those measurements.  In
\figref{sigma-beta} we plot the recovered velocity dispersion for $\beta=0$ and $\beta=1$
over a range of aperture sizes for the BC Quad with time delays and stellar
mass. The stellar half mass radius is plotted in yellow and the Einstein radius
in black. For this configuration the two radii are well separated. This plot
demonstrates that the velocity anisotropy can be inferred by combining lens
modeling and independent velocity dispersion measurements. The lens breaks
the anisotropy degeneracy at a given aperture.

>>>>

The lens equation:
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor \citep[e.g.][]{1992grle.book.....S}:
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4GD_L}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass (where $G$ is Newton's gravitational constant, and $c$ is the speed of light). For
our purposes we make use of the thin mass sheet approximation and assume that
the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These are angular diameter distances where $D_* = (c/H_0)d_*$ and $d_*$ is
a dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

@@

The trouble with solving the lens equation for the mass distribution is that the solutions are non-unique and degenerate -- many different solutions can give a statistically good reproduction of the data. To see why this is, it is instructive to simplify equation \ref{eqn:lens_equation} for a circularly symmetric lens: 
%
\begin{equation}
\vec\beta = \vec\theta - \frac{\theta_E^2}{\theta^2}\vec\theta
\label{eqn:lens_equation_circ}
\end{equation}
%
where $\theta = |\vec\theta|$, $\theta_E^2 = \frac{D_{LS}}{D_S D_L} \frac{4 GM(<\theta)}{c^2}$ is the Einstein radius, and $M(<\theta)$ is the mass enclosed within the images.

For sources on-axis (\vec\beta\ = \vec0), the images split into a degenerate Einstein ring with $\theta = \theta_E$. Even in this situation of maximal information (since the source position $\vec\beta$ cannot typically be known), there is a trivial degeneracy between $M(<\theta_E)$ and the angular diameter distances that depend on cosmological parameters (most strongly on the Hubble constant, $D_{L,S} \propto 1/H_0$). For fixed cosmology, a single source splitting into images in the most ideal situation gives us only an enclosed mass. This `steepness' degeneracy (also known as the mass-sheet degeneracy) generalises to the non-circularly symmetric case (CITES). Thus, to measure the mass profile of a lens, we require more information. This can come in the form of another source at different redshift that has its own Einstein radius, giving $M(<\theta_E')$. Or it can come in the form of additional constraints like time delay data between the images, stellar mass estimates from stellar populations synthesis modelling (CITE), stellar kinematics (CITE), or similar. However, even with further information, higher order degeneracies remain. XXX showed that there is a monopole degeneracy that allows additional mass to be deposited in-between images without any optical effect, while YYY find twisting degeneracies exist if ZZZ. These higher order degeneracies have a larger effect on shape measurement than the mass profile but remain largely uncharacterised in the literature. In \S\ref{sec:results}, we will use \Glass\ to determine how well such degeneracies can be broken by data of ever-increasing quality to measure the mass profile and shape of a lens.

@@

\subsection{A new lens modelling framework: \Glass}

\Glass\ is an evolution of the concepts employed by the free form modelling tool
\PixeLens\ but completely rewritten in Python and C. The key improvements are:
%
\begin{enumerate}
  \setcounter{enumi}{0}
  \item A modular framework allows new priors to be added and modified easily.
\end{enumerate}
%
Each prior is a simple function that adds linear constraints that operate on either
a single lens object or the entire ensemble of objects. \Glass\ comes with a number
of useful priors, but a user can write their own in the input file.
%
\begin{enumerate}
  \setcounter{enumi}{1}
  \item The basis functions approximating a model can be changed. 
\end{enumerate}
%
\Glass\ currently describes the lens mass as a collection of pixels, but the code
has been designed to support alternative methods. In particular, it is planned
to develop a module using Bessel functions. This will require a new set of 
priors that operate on these functions.
%
\begin{enumerate}
  \setcounter{enumi}{2}
  \item Non-linear constraints can be imposed in an automated post-processing step. 
\end{enumerate}
%
Once \Glass\ has generated an ensemble of models given the linear constraints, any number
of post processing functions can be applied. Not only can these functions be used to
derive new quantities from the mass models, they can also be used as a filter to 
accept or reject a model based on some non-linear constraint. The plotting functions
within \Glass\ will correctly display models that have been accepted or rejected.
%
\begin{enumerate}
  \setcounter{enumi}{3}
  \item The central region can be have a higher resolution to capture steep models. 
\end{enumerate}
%
By default, the mass distribution of the lens is described by uniform grid. However,
in the central region of a lensing galaxy where the mass profile may rise steeply,
a higher resolution can be specified that only applies to the inner pixels. These
pixels are subdivided into a smaller grid.
%
\begin{enumerate}
  \setcounter{enumi}{4}
  \item Stellar density can be used as an additional constraint.  
\end{enumerate}
%
The mass in inner regions of galaxies is often dominated by the stellar component
which one can estimate using standard mass-to-light models. This data can be added
to the potential as described later in \secref{stellar mass}. By using the stellar
mass one can place a lower bound on the mass and help constrain the inner most
mass profile.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item Point or extended mass objects can be placed in the field.
\end{enumerate}
%
\PixeLens only allows for a shear term to be added to the potential (as shown
later in \eqnref{shear}) to account for mass external to model region. This
is useful to capture the gross effects of a distant neighbour. \Glass\ also
has a shear term but additional allows for additional analytic potentials to
be included. This can be used to model substructure or multiple neighbors close
to the main lens. The substructure may have only a small effect if the lens is
a single galaxy, but if the lens is group then a potential can be added for
each of the galaxies.
%
\begin{enumerate}
  \setcounter{enumi}{5}
  \item A new uniform sampling algorithm for high dimensional spaces.
\end{enumerate}
%
At the heart of \Glass\ lies a new algorithm for sampling the high dimensional
linear space that represents the modelling solution space. This algorithm was
described and tested in \cite{}. \Glass\ is also multi-threaded allowing it to
run efficiently on many-cored machines.  The software is freely available for
download. One interesting aspect is that the input files are themselves Python
programs. This allows a large amount of sophistication in setting up a lens at
runtime and also allows \Glass\ to be used as an external library from another
user written program.

@@

In addition to the lens model aspects, \Glass\ can also run post processing
routines to analyze each model as it is produced. We have developed a number of
routines for fitting analytic models to the lensing mass profiles. One in
particular deprojects the 2d profile assuming a Hernquist profile for the
light. By fixing the velocity anisotropy $\beta$ we can determine the velocity
dispersion of the model. Of course, the true velocity dispersion is known for
the mock galaxies and we can compare our method to those measurements.  In
\figref{sigma-beta} we plot the recovered velocity dispersion for $\beta=0$ and $\beta=1$
over a range of aperture sizes for the BC Quad with time delays and stellar
mass. The stellar half mass radius is plotted in yellow and the Einstein radius
in black. For this configuration the two radii are well separated. This plot
demonstrates that the velocity anisotropy can be inferred by combining lens
modelling and independent velocity dispersion measurements. The lens breaks
the anisotropy degeneracy at a given aperture.
